
---
title: "COVID-19 Vaccination Analysis Project"
author: "Moses Merugu"
output: pdf_document
---

# Project Overview
This project aims to analyze COVID-19 vaccination data to understand vaccination trends and coverage. The objective is to provide insights into the progress of the global vaccination campaign.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Library Descriptions
```{r libraries}
# Set working directory and load necessary libraries
setwd("C:/Users/mmeru/Downloads/vaccination-analysis-project-main")
library(ggplot2) # Used for creating visualizations
library(formatR) # Assists in formatting R code
library(tidyverse) # Collection of R packages for data manipulation
```

# Data Loading and Preprocessing
```{r data}

# Loading COVID-19 vaccination data
vax <- read_csv("https://raw.githubusercontent.com/govex/COVID-19/master/data_tables/vaccine_data/global_data/time_series_covid19_vaccine_doses_admin_global.csv")

# Loading COVID-19 hospital bed data
beds <- read_csv("hospitalbed.csv")

# Loading COVID-19 demographics data
demo <- read_csv("demographics.csv")
```

# Clean vaccine data
```{r vaccines, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
# VACCINE DATA
# Filter out provinces and countries with no population
vax <- vax %>% filter(is.na(Province_State), !is.na(Population)) %>% view()
# Tidy number of vaccinations on given date
vax <- vax %>% pivot_longer(-c(1:12), names_to = "Date", values_to = "Vaccinations", values_drop_na = TRUE) %>% view()
# Delete irrelevant columns
vax <- vax[,-c(1,2,3,4,5,6,7,9,10,11)] %>% view()
# Filter out rows containing dates with 0 vaccinations
vax <- vax %>% filter(!Vaccinations == 0) %>% view()
# Calculate vaccination rate and add respective column
vax <- vax %>% select(Country_Region, Population, Vaccinations) %>% group_by(Country_Region) %>% mutate(Vaccination_Rate = Vaccinations / Population) %>% view()
# Add column that tracks days since first vaccination
vax <- vax %>% group_by(Country_Region) %>% mutate(Days_Since_First_Vaccination = 1:n()) %>% view()
```

# Clean hospital bed data
```{r beds, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
# BEDS DATA
# Most recent year appears first, keep the first bed value per country using summarize()
# Year column is not needed
beds <- beds %>% group_by(Country) %>% summarize(Beds=first(`Hospital beds (per 10 000 population)`)) %>% view()
```

# Clean demographics data
```{r demos, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
# DEMOGRAPHICS DATA
# Tidy data
demo <- demo %>% pivot_wider(id_cols = -'Series Name', names_from = 'Series Code', values_from = YR2015) %>% view()
# Add male and female data together
demo <- demo %>% mutate(SP.POP.0014.IN=SP.POP.0014.MA.IN+SP.POP.0014.FE.IN) %>% mutate(SP.POP.80UP=SP.POP.80UP.FE+SP.POP.80UP.MA) %>% mutate(SP.POP.1564.IN=SP.POP.1564.MA.IN+SP.POP.1564.FE.IN) %>% mutate(SP.DYN.AMRT=SP.DYN.AMRT.MA+SP.DYN.AMRT.FE) %>% mutate(SP.POP.TOTL.IN=SP.POP.TOTL.FE.IN+SP.POP.TOTL.MA.IN) %>% mutate(SP.POP.65UP.IN=SP.POP.65UP.FE.IN+SP.POP.65UP.MA.IN) %>% view()
# Drop country code and gender specific columns, filter NAs
demo <- demo[,-c(2,6:17)] %>% filter(!is.na(SP.DYN.LE00.IN), !is.na(SP.URB.TOTL), !is.na(SP.POP.0014.IN), !is.na(SP.POP.80UP), !is.na(SP.POP.1564.IN), !is.na(SP.DYN.AMRT), !is.na(SP.POP.TOTL.IN), !is.na(SP.POP.65UP.IN) ) %>% view()
```

# Format data
```{r format, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
# UNIFORMING COUNTRY NAMES TO MATCH VACCINE DATA
beds <- beds %>% mutate(Country = replace(Country, Country == "Iran (Islamic Republic of)", "Iran"))
beds <- beds %>% mutate(Country = replace(Country, Country == "Republic of Korea", "South Korea"))
beds <- beds %>% mutate(Country = replace(Country, Country == "United Kingdom of Great Britain and Northern Ireland", "United Kingdom"))
beds <- beds %>% mutate(Country = replace(Country, Country == "Bolivia (Plurinational State of)", "Bolivia"))
beds <- beds %>% mutate(Country = replace(Country, Country == "Lao People's Democratic Republic", "Laos"))
beds <- beds %>% mutate(Country = replace(Country, Country == "Venezuela (Bolivarian Republic of)", "Venezuela"))
beds <- beds %>% mutate(Country = replace(Country, Country == "Republic of Moldova", "Moldova"))
beds <- beds %>% mutate(Country = replace(Country, Country == "United States of America", "US"))
beds <- beds %>% mutate(Country = replace(Country, Country == "Viet Nam", "Vietnam"))

demo <- demo %>% mutate(`Country Name` = replace(`Country Name`, `Country Name` == "Korea, Rep.", "South Korea"))
demo <- demo %>% mutate(`Country Name` = replace(`Country Name`, `Country Name` == "Iran, Islamic Rep.", "Iran"))
demo <- demo %>% mutate(`Country Name` = replace(`Country Name`, `Country Name` == "Venezuela, RB", "Venezuela"))
demo <- demo %>% mutate(`Country Name` = replace(`Country Name`, `Country Name` == "St. Vincent and the Grenadines", "Saint Vincent and the Grenadines"))
demo <- demo %>% mutate(`Country Name` = replace(`Country Name`, `Country Name` == "St. Lucia", "Saint Lucia"))
demo <- demo %>% mutate(`Country Name` = replace(`Country Name`, `Country Name` == "Slovak Republic", "Slovakia"))
demo <- demo %>% mutate(`Country Name` = replace(`Country Name`, `Country Name` == "Czech Republic", "Czechia"))
demo <- demo %>% mutate(`Country Name` = replace(`Country Name`, `Country Name` == "Bahamas, The", "Bahamas"))
demo <- demo %>% mutate(`Country Name` = replace(`Country Name`, `Country Name` == "United States", "US"))
```

# Merging tables
```{r merge, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
# Perform inner joins to merge tables
join <- beds %>% inner_join(vax, by=c(Country="Country_Region")) %>% inner_join(demo, by=c(Country="Country Name")) %>% view()

# Rearrange column order to match example
final <- join[,c(1,5,4,3,6,2,7,8)]

view(final)
```

# Plots and Linear Models
```{r plots, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
# PLOTS AND LINEAR MODELS
# Scatterplot of only the most recent vaccination rate for every country and the number of days since first vaccination
forplot <- final %>% relocate(Vaccinations, .after = Country) %>% relocate(Vaccination_Rate, .after = Country)
scatter <- ggplot(data=forplot) + geom_point(mapping=aes(x=Days_Since_First_Vaccination, y=Vaccination_Rate))
scatter

m1 <- lm(data = final, Vaccination_Rate ~ Days_Since_First_Vaccination)
summary(m1) # R-squared: 0.6125

m2 <- lm(data = final, Vaccination_Rate ~ Days_Since_First_Vaccination + Beds)
summary(m2) # R-squared: 0.6341

m3 <- lm(data = final, Vaccination_Rate ~ Days_Since_First_Vaccination + SP.DYN.LE00.IN)
summary(m3) # R-squared: 0.7473

m4 <- lm(data = final, Vaccination_Rate ~ Days_Since_First_Vaccination + SP.URB.TOTL)
summary(m4) # R-squared: 0.6131

m5 <- lm(data = final, Vaccination_Rate ~ Days_Since_First_Vaccination + SP.URB.TOTL + SP.DYN.LE00.IN)
summary(m5) # R-squared: 0.7478

# Organize 5 models and corresponding R2 values into data frame
df <- data.frame(Model=c("M1", "M2", "M3", "M4", "M5"),
R2=c(summary(m1)$r.squared, summary(m2)$adj.r.squared, summary(m3)$adj.r.squared, summary(m4)$adj.r.squared, summary(m5)$adj.r.squared))

# Create bar plot comparing models and their R2 values
bar <-ggplot(data=df, aes(x=Model, y=R2)) + geom_bar(stat="identity")
bar

```